Predicting Probability of Participants' Vote of Conservative and Liberty in 2025 Canadian election 

### 2024-04-19 paper/paper(1).qmd
Add files via upload

To enhance control and prevention of illegal immigration, the Liberal Party has suggested modifications to the Immigration and Refugee Protection Act.
They propose establishing an annual refugee quota of 250, specifically aimed at supporting human rights defenders, journalists, and humanitarian workers seeking safety and resettlement in Canada.
Looking ahead, the Liberal Party intends to incrementally increase this refugee quota, aiming for refugees to comprise 15% of all immigrants.
Conversely, the share of economic immigrants is planned to decrease to 58%, while the family reunification immigrant ratio will remain constant.
These proportions are expected to be maintained over the next two years.
Moreover, the Canadian national government recently unveiled an ambitious plan to admit 500,000 migrants annually by 2025, looking for nearly 1.5 million newcomers in Canada within the next three years (Canada's Immigration Levels Plan 2024-2026).
Historically, the Liberal government's immigration strategy tends to be more permissive, offering broader entry routes compared to that of the Conservative Party[@Nakhaie_2006].
Based on our analysis of historical trends, governmental policies, and electoral procedures, our hypothesis aligns with the prevailing expectation that the Conservatives will emerge as the likely winners in the 2025 election.
**We anticipate that the Liberals will likely secure the highest number of votes, with the Conservatives expected to closely follow in second place.** Our research indicates that the vote-counting process in Canada significantly influences the outcome, as voters do not directly elect a specific prime minister candidate.
Initially, all votes are tallied based on first preferences.
If a candidate obtains 50% of the vote, they win.
If not, the candidate with the fewest votes is eliminated[@Elections_2023].
Then, votes from supporters of the eliminated candidate are redistributed to their second-choice candidate.
This iterative redistribution continues until one candidate achieves 50% of the vote.
Hence, our objective is to forecast the election probabilities for the Conservative and Liberal parties and compare which is higher[@Elections_2023].

@@

In this paper, I used binary logistic regression models to forecast the likelihood of individuals affiliating with the Conservative or Liberal parties.Binary logistic regression is a statistical technique tailored for predicting binary outcomes, where the variable under consideration has two possible results.
In our study, these outcomes represent the probability of individuals aligning with each of the two parties.Each party's likelihood will be predicted through separate logistic regression models, with the response variables indicating whether individuals vote for the corresponding party or not.
The predictors used in these models include age, province, education, religion, and income before tax.
The variable "sex" will be initially included but later removed after AIC model testing, as further elaborated in the model selection section.
This regression approach is particularly advantageous when there's a nonlinear relationship between independent and dependent variables, especially in scenarios with categorical dependent variables like ours.
Additionally, we will incorporate post-stratification in our analysis to enhance accuracy.
Post-stratification helps refine precision estimates by ensuring that the sample is more representative of the entire population.
We utilize this technique to predict the probability of voter shares for the Liberal or Conservative parties in the upcoming election.

@@

Recent research highlights that the primary concerns for Canadian voters leading up to the 2025 election include the rising cost of living, housing, healthcare, the economy, and the environment.
These issues are pivotal to the electorate, and the political parties’ stances on them are under intense scrutiny.
Our analysis, drawn from five different polls, suggests a marginal preference for the Conservative Party over the Liberal Party.
This trend is echoed in Barik's 2023 article, which indicates a public perception of the Conservative Party's superior handling of these key issues.
Thus, our working hypothesis posits that the Conservative Party holds a slender advantage in the race for the upcoming election[@Barik_2023].

For our model development, we chose predictors from census and survey data, refined through thorough research and reflecting the public's primary concerns.
We proceeded with **binary logistic regression models**—one for each major party—to predict voter preferences based on complex social backgrounds.
After refining the models using Akaike's Information Criteria to eliminate superfluous predictors, we established age, province, religion, education, and pre-tax income as our final variables.

Post-stratification was applied to these predictors, revealing that the Conservative Party has a projected win probability of 25%, marginally outpacing the Liberal Party's 24.9%.
These projections reinforce our initial hypothesis that the Conservative Party is the slight favorite for the 2025 election, based on our analysis of the data at hand.
Our findings are in line with prior research, indicating that the main electoral contest will be between the Conservative and Liberal parties.

## Consideration of Potential Bias

@@

@@

In conclusion, our binary logistic regression approach, supplemented with post-stratification, has yielded a data-driven forecast for the 2025 Canadian Election.
Aligning with Barik’s analysis and multiple polls, our findings suggest the Conservative Party is poised for a narrow victory over the Liberal Party, informed by an intricate blend of sociopolitical factors and current Canadian sentiment[@Barik_2023].

>>>>

To enhance control and prevention of illegal immigration, the Liberal Party has suggested modifications to the Immigration and Refugee Protection Act.
They propose establishing an annual refugee quota of 250, specifically aimed at supporting human rights defenders, journalists, and humanitarian workers seeking safety and resettlement in Canada.
Looking ahead, the Liberal Party intends to incrementally increase this refugee quota, aiming for refugees to comprise 15% of all immigrants.
Conversely, the share of economic immigrants is planned to decrease to 58%, while the family reunification immigrant ratio will remain constant.
These proportions are expected to be maintained over the next two years.
Moreover, the Canadian national government recently unveiled an ambitious plan to admit 500,000 migrants annually by 2025, looking for nearly 1.5 million newcomers in Canada within the next three years (Canada's Immigration Levels Plan 2024-2026).
Historically, the Liberal government's immigration strategy tends to be more permissive, offering broader entry routes compared to that of the Conservative Party[@Nakhaie_2006].
Based on our analysis of historical trends, governmental policies, and electoral procedures, our hypothesis aligns with the prevailing expectation that the Conservatives will emerge as the likely winners in the 2025 election.
**We anticipate that the Liberals will likely secure the highest number of votes, with the Conservatives expected to closely follow in second place.** Our research indicates that the vote-counting process in Canada significantly influences the outcome, as voters do not directly elect a specific prime minister candidate.
Initially, all votes are tallied based on first preferences.
If a candidate obtains 50% of the vote, they win.
If not, the candidate with the fewest votes is eliminated[@Elections_2023].
Then, votes from supporters of the eliminated candidate are redistributed to their second-choice candidate.
This iterative redistribution continues until one candidate achieves 50% of the vote.
Hence, our objective is to forecast the election probabilities for the Conservative and Liberal parties and compare which is higher[@Elections_2023]. The estimand of our research is relationship between age, gender, province, religion, income and education and probability of log odds of vote Conservative or Liberal party through binary logistic regression model.

@@

In this paper, I used binary logistic regression models to forecast the likelihood of individuals affiliating with the Conservative or Liberal parties.Binary logistic regression is a statistical technique tailored for predicting binary outcomes, where the variable under consideration has two possible results.
In our study, these outcomes represent the probability of individuals aligning with each of the two parties.Each party's likelihood will be predicted through separate logistic regression models, with the response variables indicating whether individuals vote for the corresponding party or not.
The predictors used in these models include age, gender, province, education, religion, and income before tax.
This regression approach is particularly advantageous when there's a nonlinear relationship between independent and dependent variables, especially in scenarios with categorical dependent variables like ours.
Additionally, we will incorporate post-stratification in our analysis to enhance accuracy.
Post-stratification helps refine precision estimates by ensuring that the sample is more representative of the entire population.
We utilize this technique to predict the probability of voter shares for the Liberal or Conservative parties in the upcoming election.

@@

Founded at the inception of Canada in 1867, the Liberal Party has long been a prominent player in the nation's political arena. Rooted in the principles of liberalism—a cornerstone of Canadian political ideology—the party champions a balanced fiscal approach, advocating for judicious tax reductions to foster economic expansion, boost investment, and invigorate consumer spending. At the heart of the Liberal Party's platform is a commitment to economic vitality and job creation, with a focus on propelling Canada's growth and ensuring ample employment opportunities. The party also prioritizes the well-being of its constituents, pledging to sustain and enhance public services including healthcare, pensions, and education. Committed to environmental stewardship, the Liberal Party envisions a greener Canada, aiming for reduced carbon emissions and heightened conservation efforts.
Our analysis, drawn from five different polls, suggests a marginal preference for the Liberal Party over the Conservative Party.
This trend is echoed in Russell's 2023 article, which indicates a public perception of the Liberal Party's superior handling of these key issues.
Thus, our working hypothesis posits that the Liberal Party holds a slender advantage in the race for the upcoming election[@Brussell2023two].

For our model development, we chose predictors from census and survey data, refined through thorough research and reflecting the public's primary concerns.
We proceeded with **binary logistic regression models**—one for each major party—to predict voter preferences based on complex social backgrounds.
After refining the models using Akaike's Information Criteria to eliminate superfluous predictors, we established age, gender, province, religion, education, and pre-tax income as our final variables.

Post-stratification was applied to these predictors, revealing that the Liberal Party has a projected probability of 28.83%, marginally outpacing the Conservative Party's 28.66%.
These projections reinforce our initial hypothesis that the Liberal Party is the slight favorite for the 2025 election, based on our analysis of the data at hand.
Our findings are in line with prior research, indicating that the main electoral contest will be between the Conservative and Liberal parties.

## Consideration of Potential Bias and Limitations

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
step_lib_counts <- cleaned_survey_data %>%
  group_by(age, sex, province, education, religion, income_before_tax) %>%
  summarise(N_j=n(), .groups = 'drop')

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
step_lib_counts <- cleaned_survey_data %>%
  group_by(age, sex, province, education, religion, income_before_tax) %>%
  summarise(N_j=n(), .groups = 'drop')

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide', include =FALSE}

lib_predictions <- predict(step_lib, newdata = step_lib_counts, type = "response")

lib_CI <- data.frame(
  estimate = lib_predictions,

  observation = seq_along(lib_predictions)
)

head(lib_CI)

```{r}
#| label: fig-liberal-votes-and-consrvative_votes
#| fig-cap: "predicted probabilities for the Conservative and Liberal parties"
#| echo: false
cons_probs <- predict(step_cons, newdata = cleaned_survey_data, type = "response")
lib_probs <- predict(step_lib, newdata = cleaned_survey_data, type = "response")

combined_predictions <- data.frame(
  PredictedProbability = c(cons_probs, lib_probs),
  Party = rep(c("Conservative", "Liberal"), each = length(cons_probs))
)

ggplot(combined_predictions, aes(x = PredictedProbability, fill = Party)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Party, nrow = 2) +
  scale_fill_manual(values = c("red", "green")) +
  labs(x = "Predicted Probability", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

@fig-liberal-votes-and-consrvative_votes provide density plots illustrating the distribution of predicted probabilities for the Conservative and Liberal parties, respectively. Both plots reveal a slight left skew, indicating a greater concentration of predictions at the lower probability range, nearer to 0.2. This skew suggests that while both parties are predicted to have a significant chance of receiving votes, there's a tendency for the model to assign a moderate likelihood rather than a strong certainty of voting outcomes.

Notably, the Liberal party's density plot reaches a higher peak, which points to a narrower, more defined range of predicted probabilities. This contrast implies that the model predicts voting probabilities for the Liberal party with more confidence, as the predictions are less spread out over a range of values compared to the Conservative party.

Additionally, the visible overlap between the two distributions signifies a degree of ambiguity or uncertainty in differentiating the predicted probabilities for the two parties. The shared area of the plots indicates that for a substantial portion of predictions, the likelihood of individuals voting for either party is comparably uncertain.

The nearly identical yPS values for the Conservative and Liberal parties, standing at 0.2865 and 0.2883 respectively, suggest a possibility of incomplete or inaccurate data processing. Such closeness in predictive scores could point to calculation errors during the yPS estimation, which in turn could impact the determination of the potential election winner.

@@

In conclusion, our binary logistic regression approach, supplemented with post-stratification, has yielded a data-driven forecast for the 2025 Canadian Election.
Aligning with Russell’s analysis and multiple polls, our findings suggest the Liberal Party is poised for a narrow victory over the Conservative Party, informed by an intricate blend of sociopolitical factors and current Canadian sentiment[@russell2023two].
